A physical foundation of Differential Topology

### 2023-05-19 Articles/2023-PhysicalDifferentialTopology/PhysicalDifferentialTopology.tex
Minor fixes

In differential topology, one defines local objects (e.g. vectors and forms) and then defines finite quantities associated to regions through integration. For example, one defines volume forms and then defines its integral over a region. Physics proceeds the opposite way. Experimentally, we do not measure directly a mass density, but rather we measure mass and volume of a finite region and take the ratio. We then imagine to take smaller and smaller regions, and the mass density is ratio in the limit.\footnote{The Radon–Nikodym derivative, for example, follows this approach.} Ideally, our definitions should mirror this process to which there are at least to major advantages. First, it is a lot easier to build physical intuition on the finite objects than on the local ones. Second, we can understand more clearly what physical requirements and idealizations are embedded in the definitions. For example, one basic assumption to define the mass density is that if we partition the volume, the total mass is the sum of the one found in each part. Given the equivalence of mass and energy, this is strictly true only if there is no interaction energy between the parts. Exactly what happens is those cases, then, is not clear, and starting directly with local objects may hide actual conceptual problems.

@@

While units and physical dimensions play a crucial role in physics, these ignored by the mathematics with important consequences. First of all, most people, physicists and mathematicians alike, miss that transformation rules of tensor components (i.e. covariance and contravariance) are primarily about units. That is, some quantities are expressed in base units (e.g. position and velocity) and others in derived units (e.g. velocity), therefore a change of definition in the first induces a change in the others. Since mathematics does not explicitly formalize units, it can do two things: either just define objects based on their transformation rules or look for objects that, by definition, follow the same transformation rules.\footnote{It seems that,in the mathematical world, the latter approach has been gaining favor with respect to the second.} To give an example of why this is problematic, in differential geometry one talks about tangent and cotangent bundle. Roughly speaking, the first pairs a point with a vector and the second pairs a point with a one form. It is often said that Lagrangian mechanics works on the tangent bundle, while Hamiltonian mechanics work on the cotangent bundle since one identifies velocity with a vector and momentum with a one form. Conceptually, however, the space in both cases represents all possible states of point particles. It is a manifold and the state is identified by a set of six variables. The fact that we choose position/velocity or position/momentum does not change the nature of the state space. Moreover, nothing prevents us to use units of velocity that are different from the one derived from the units of position.\footnote{This is actually often done for polar and cylindrical coordinates. The basis induced by the coordinates is not orthonormal, and is often replaced by one that is. It is not rare for people to not be aware of this and getting confused.} Therefore the notion of tangent/cotangent bundle is there to capture unit relationships between state variables. We can now ask: are there similar relationships that we may want to capture? If we have multiple particles, we will typically want to express their position in the same units. If we have the state space for $N$ particles, then, the transformation rules would be more restrictive than those of a tangent or cotangent bundle.

>>>>

In differential topology, one defines local objects (e.g. vectors and forms) and then defines finite quantities associated to regions through integration. For example, one defines volume forms and then defines its integral over a region. Physics proceeds the opposite way. Experimentally, we do not measure directly a mass density, but rather we measure mass and volume of a finite region and take the ratio. We then imagine to take smaller and smaller regions, and the mass density is ratio in the limit.\footnote{The Radon–Nikodym derivative, for example, follows this approach.} This is similar for other quantities: we do not measure the magnetic field at a point but rather the magnetic flux over an ever smaller finite surface; we do not measure the force at a point but rather the work done by a force over an ever smaller finite path. Ideally, our definitions should mirror this process, which would have at least two major advantages. First, it is a lot easier to build physical intuition on the finite objects than on the local ones. Second, we can understand more clearly what physical requirements and idealizations are embedded in the definitions. For example, one basic assumption to define the mass density is that if we partition the volume, the mass in the whole volume is the sum of the one found in each part. Given the equivalence of mass and energy, this is strictly true only if there is no interaction energy between the parts. Exactly what happens is those cases, then, is not clear, and starting directly with local objects may hide actual conceptual problems.

@@

While units and physical dimensions play a crucial role in physics, these ignored by the mathematics with important consequences. First of all, most people, physicists and mathematicians alike, miss that transformation rules of tensor components (i.e. covariance and contravariance) are primarily about units. That is, some quantities are expressed in base units (e.g. position and velocity) and others in derived units (e.g. velocity), therefore a change of definition in the first induces a change in the others.\footnote{As a more concrete example, consider the expression for work $W = \int_\lambda f_i dx^i$ as the line integral of the force. Work is an invariant meaning that its unit is independent of the the units used for the coordinates $x^i$. However, the units used for the components of the force will necessarily be the ration between the work unit and the corresponding unit for the coordinate.} Since mathematics does not explicitly formalize units, it can do two things: either just define objects based on their transformation rules or look for objects that, by definition, follow the same transformation rules.\footnote{It seems that,in the mathematical world, the latter approach has been gaining favor with respect to the second.} To give an example of why this is problematic, in differential geometry one talks about tangent and cotangent bundle. Roughly speaking, the first pairs a point with a vector and the second pairs a point with a one form. It is often said that Lagrangian mechanics works on the tangent bundle, while Hamiltonian mechanics work on the cotangent bundle since one identifies velocity with a vector and momentum with a one form. Conceptually, however, the space in both cases represents all possible states of point particles. It is a manifold and the state is identified by a set of six variables. The fact that we choose position/velocity or position/momentum does not change the nature of the state space. Moreover, nothing prevents us to use units of velocity that are different from the one derived from the units of position.\footnote{This is actually often done for polar and cylindrical coordinates. The basis induced by the coordinates is not orthonormal, and is often replaced by one that is. It is not rare for people to not be aware of this and getting confused.} Therefore the notion of tangent/cotangent bundle is there to capture unit relationships between state variables. We can now ask: are there similar relationships that we may want to capture? If we have multiple particles, we will typically want to express their position in the same units. If we have the state space for $N$ particles, then, the transformation rules would be more restrictive than those of a tangent or cotangent bundle.
